---
title: "Wildfire Exploration"
author: "Iris Foxfoot"
date: "2/3/2022"
output: 
  html_document:
    toc: true # table of content true
    toc_depth: 3  # upto three depths of headings (specified by #, ## and ###)
    number_sections: true  ## if you want number sections at each table header
    theme: united  # many options for theme, this one is my favorite.
    highlight: tango  # specifies the syntax highlighting style
    toc_float: true #makes table of contents float while scrolling
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
library(tidyverse)
library(here)
```

# Introduction

What is the purpose and objectives of this project

## Background

fire situation in North America, climate change, etc

## Why this model is useful

What will it be used for?

## Data and Packages Used

where the data came from, what it contains, etc

```{r}
#read in datasets
ca_wildfire <- read_csv(here("California_Fire_Incidents.csv"))

us_wildfire <- read_csv(here("archive", "FW_Veg_Rem_Combined.csv"))
```

```{r}
#explore us wildfire dataset

#17630 rows with no NAs
us_wildfire_no_na <- na.omit(us_wildfire)

#41132 entries with weather data
us_wildfire_weather <- us_wildfire %>% 
  filter(weather_file != "File Not Found")

#note, put out time is not accurate but can be recalculated

#histogram of temp 30 days before fire
hist(us_wildfire_weather$Temp_pre_30)

hist(us_wildfire_weather$Temp_pre_7)
```

# Methods

Overview of methods

## Cleaning Data
 
what we did to clean the data

## Exploratory Analysis

~ a few key graphs, exploring data

ideas for graphs
  - map of fires
  - causes of fires
  - duration of fires
  - vegetation
  - destruction
  - weather conditions

## Splitting Data

Split into testing and training data


## Modeling

what type of modeling will we do? we must try more than one type

## Cross Validation

Which model is best?

## Final Model Selection

## Test Error

# Results

predictions without climate change

How do predictions change with 2.5C increase in temp?

# Conclusion


